\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{geometry}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{caption}

\geometry{margin=1in}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstset{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\small,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2,
    language=C++
}

\title{\textbf{Numerical Methods From Scratch: Engineering Mathematics Implemented in Modern C++}}
\author{Gemini CLI Engineering Team}
\date{June 2, 2026}

\begin{document}

\maketitle

\begin{abstract}
This report details the design and implementation of a professional scientific computing library built from first principles using the C++20 standard. The project, \textit{Numerical Methods From Scratch}, emphasizes mathematical rigor, numerical stability, and computational efficiency. We cover a broad spectrum of numerical analysis, including root-finding, linear algebra, interpolation, calculus, and differential equations. The implementation leverages modern C++ features, RAII, and template metaprogramming to provide a robust framework for engineering and mathematical computations.
\end{abstract}

\newpage
\tableofcontents
\newpage

\section{Introduction}
The field of numerical analysis is foundational to modern engineering and scientific discovery. While high-level environments like MATLAB or Python's NumPy provide powerful abstraction, there is significant value in implementing core algorithms from scratch to understand their limitations, stability characteristics, and performance trade-offs.

This project aims to bridge the gap between abstract mathematical theory and concrete software implementation. By using C++20, we take advantage of advanced features such as concepts, ranges, and improved template support to create a type-safe and performant library. The primary dependencies include \texttt{Eigen} for matrix operations, \texttt{matplotplusplus} for visualization, and \texttt{GoogleTest} for rigorous verification.

\section{Root Finding Methods}
Root-finding involves finding $x$ such that $f(x) = 0$. We implemented four fundamental methods.

\subsection{Bisection Method}
The Bisection Method relies on the Intermediate Value Theorem. For a continuous function $f$, if $f(a)f(b) < 0$, a root exists in $[a, b]$.
\begin{equation}
    c = \frac{a+b}{2}
\end{equation}
While slow (linear convergence), it is unconditionally stable.

\subsection{Newton-Raphson Method}
Newton's method uses the first-order Taylor expansion:
\begin{equation}
    x_{n+1} = x_n - \frac{f(x_n)}{f'(x_n)}
\end{equation}
It offers quadratic convergence ($e_{n+1} \approx C e_n^2$) but requires the existence and calculation of the derivative.

\subsection{Secant Method}
The Secant method approximates the derivative using two points:
\begin{equation}
    x_{n+1} = x_n - f(x_n) \frac{x_n - x_{n-1}}{f(x_n) - f(x_{n-1})}
\end{equation}
It avoids analytical derivatives while maintaining superlinear convergence ($\phi \approx 1.618$).

\section{Linear Algebra}
Solving $Ax=b$ is the workhorse of scientific computing.

\subsection{LU Decomposition}
We implemented Doolittle's algorithm to factor $A = LU$, where $L$ is unit lower triangular and $U$ is upper triangular.
\begin{equation}
    A = \begin{pmatrix} 1 & 0 \\ l_{21} & 1 \end{pmatrix} \begin{pmatrix} u_{11} & u_{12} \\ 0 & u_{22} \end{pmatrix}
\end{equation}
This allows solving multiple systems with different $b$ in $O(n^2)$ time after an initial $O(n^3)$ decomposition.

\subsection{Iterative Solvers}
For large sparse systems, we implemented Jacobi and Gauss-Seidel methods. Gauss-Seidel uses the iteration:
\begin{equation}
    x_i^{(k+1)} = \frac{1}{a_{ii}} \left( b_i - \sum_{j < i} a_{ij} x_j^{(k+1)} - \sum_{j > i} a_{ij} x_j^{(k)} \right)
\end{equation}
Convergence is guaranteed for strictly diagonally dominant matrices.

\section{Interpolation}
Interpolation fits a function to discrete data points.

\subsection{Cubic Spline Interpolation}
Cubic splines provide a piecewise cubic polynomial $S(x)$ that is $C^2$ continuous.
\begin{equation}
    S_i(x) = a_i + b_i(x-x_i) + c_i(x-x_i)^2 + d_i(x-x_i)^3
\end{equation}
Unlike high-degree global polynomials, splines avoid Runge's phenomenon and provide smooth approximations.

\section{Numerical Differentiation}
We utilize finite difference quotients to approximate derivatives.
\begin{itemize}
    \item \textbf{Central Difference}: $f'(x) \approx \frac{f(x+h) - f(x-h)}{2h} + O(h^2)$
\end{itemize}
The central difference is preferred for its second-order accuracy.

\section{Numerical Integration}
Quadrature rules approximate $\int_a^b f(x) dx$.

\subsection{Gaussian Quadrature}
Gaussian quadrature selects optimal nodes $x_i$ and weights $w_i$:
\begin{equation}
    \int_{-1}^1 f(x) dx \approx \sum_{i=1}^n w_i f(x_i)
\end{equation}
An $n$-point rule is exact for polynomials of degree $2n-1$.

\section{Ordinary Differential Equations}
We solve Initial Value Problems (IVPs) of the form $y' = f(t, y)$.

\subsection{Runge-Kutta 4 (RK4)}
RK4 is a high-accuracy fourth-order method:
\begin{align}
    k_1 &= f(t_n, y_n) \not b\\
    k_2 &= f(t_n + h/2, y_n + h k_1/2) \\
    k_3 &= f(t_n + h/2, y_n + h k_2/2) \\
    k_4 &= f(t_n + h, y_n + h k_3) \\
    y_{n+1} &= y_n + \frac{h}{6}(k_1 + 2k_2 + 2k_3 + k_4)
\end{align}

\section{Error Analysis}
Numerical errors are classified into:
\begin{itemize}
    \item \textbf{Truncation Error}: From approximating an infinite process with a finite one.
    \item \textbf{Round-off Error}: From finite floating-point precision.
\end{itemize}
We utilize $Result<T>$ structures to track the convergence history and residual errors of every computation.

\section{Results and Visualizations}
We generated several plots to validate our implementations. (Note: Plots reference the \texttt{assets/images/} directory).

\subsection{Convergence of Root Finding}
The comparison between Bisection and Newton-Raphson highlights the quadratic nature of Newton's method. While Bisection requires $\sim 40$ iterations for $10^{-12}$ precision, Newton's method converges in fewer than 10.

\subsection{ODE Solver Accuracy}
Plotting the solution of $y' = y$ demonstrates that RK4 remains highly accurate even with relatively large step sizes $h$, whereas the Euler method accumulates significant local truncation error.

\section{Conclusion}
The \textit{Numerical Methods From Scratch} library demonstrates that modern C++20 is an excellent choice for scientific computing. By combining mathematical rigor with robust engineering practices, we have created a library that is both educationally valuable and computationally sound. Future work could include the implementation of implicit ODE solvers for stiff systems and multi-dimensional root-finding.

\newpage
\begin{thebibliography}{9}
\bibitem{burden}
Burden, R. L., \& Faires, J. D. (2010). \textit{Numerical Analysis}. Cengage Learning.
\bibitem{knuth}
Knuth, D. E. (1997). \textit{The Art of Computer Programming, Volume 2: Seminumerical Algorithms}. Addison-Wesley.
\bibitem{cpp20}
ISO/IEC. (2020). \textit{ISO/IEC 14882:2020 Programming languages --- C++}.
\end{thebibliography}

\end{document}
